### 2018-04-18 1_intro.tex
Cleanup intro and wmm chapters.

Traditionally, studies related to concurrent programming concern on more fundamental theoretical questions of designing race-free and lock-free parallel algorithms, asynchronous data structures and synchronisation primitives of a programming language. Unfortunately,
when it comes to 
%in cases of 
the real-world concurrent programs, the algorithmic level of abstraction is not enough for guaranteeing their properties of correctness and reliability. The reasons of this fact lie in the code optimisations that both compiler and hardware perform in order to increase performance as much as possible. For instance, Figure~\ref{simple_wmm_x86} provides simple example of
unexpected state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' reachable on x86 machines (such little examples that illustrate specific behaviour of a WMM are called \textit{litmus tests}).
%reordering of memory access instructions within single process allowed by the x86-TSO weak memory model, which potentially breaks the program logic. 
This state is allowed because in x86 architecture each processor may cache the write to shared memory variable into its local write buffer, so that they do not become visible by other processes immediately. In the example, the write `\texttt{MOV~[x],1}' performed by process \texttt{P0} stores value~\texttt{1} to the shared variable~\texttt{[x]} into the write buffer of process~\texttt{P0}. Meanwhile, the write cache of the process \texttt{P1} may not have updated version of the variable~\texttt{[x]}, neither may have the main memory, so that the read `\texttt{MOV~EBX,[x]}' performed in the process~\texttt{P1} may read the initial value~\texttt{0}, even if this variable has been already updated in another thread.
%More presice description of x86-TSO memory model is given in Chapter~\ref{ch:wmm:x86}.

@@

@@

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{\url{diy.inria.fr}} is a sofware suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy7}, \tool{diycross7} and \tool{diyone7}, the litmus tests concrete executor \tool{litmus7} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd7} that implements reachability analysis for exploring states reachable under specified~WMM.

All the \tool{diy} tools work only with single memory model, however, in real life we face serious engineering problems involving necessity to model more than one execution environment. One of these problems is the \textit{portability} of the program from one hardware architecture to another. A program written in a high-level language is then compiled for different hardware. Even if all the compiler optimisations were disabled (which is rare case nowadays), the behaviour of two compiled versions of the same program may differ due to differences between hardware memory models. As the result, the program compiled for the platform $X$, can reach states that are unreachable on the platform $Y$, or some states that are reachable on $Y$ can become unreachable on $X$. We declare these cases as \textit{portability bugs}.

>>>>

Traditionally, studies related to concurrent programming concern on more fundamental theoretical questions of designing race-free and lock-free parallel algorithms, asynchronous data structures and synchronisation primitives of a programming language. Unfortunately, when it comes to the real-world concurrent programs, the algorithmic level of abstraction is not enough for guaranteeing their properties of correctness and reliability. The reasons of this fact lie in the code optimisations that both compiler and hardware perform in order to increase performance as much as possible. For instance, Figure~\ref{simple_wmm_x86} provides simple example of unexpected state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' reachable on x86 machines (such little examples that illustrate specific behaviour of a WMM are called \textit{litmus tests}).
%reordering of memory access instructions within single process allowed by the x86-TSO weak memory model, which potentially breaks the program logic. 
This state is allowed because in x86 architecture each processor may cache the write to shared memory variable into its local write buffer, so that they do not become visible by other processes immediately. In the example, the write `\texttt{MOV~[x],1}' performed by process \texttt{P0} stores value~\texttt{1} to the shared variable~\texttt{[x]} into the write buffer of process~\texttt{P0}. Meanwhile, the write cache of the process \texttt{P1} may not have updated version of the variable~\texttt{[x]}, neither may have the main memory, so that the read `\texttt{MOV~EBX,[x]}' performed in the process~\texttt{P1} may read the initial value~\texttt{0} even if this variable has been already updated in another thread. These problems have lead to the need for formalisation of semantics of memory operations within different concurrent architectures defined by \textit{weak	memory models (WMM)}.

@@

Research of weak memory models firstly aimes to \textit{formalise} develop the formal approach of understanding programs with respect to weak memory models which is systematic, sound and complete. The first (and so far the only) such a framework was presented in 2010~\cite{alglave2010shared}.
%
Secondly, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, that are written in natural language and thus suffer from ambiguities and incompletenesses. Over last decade the memory models have been extracted for most mainstream multiprocessor architectures, such as x86-TSO (\textit{Total Store Order}) model for x86 architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding}, for Alpha~\cite{?} and Sparc~\cite{?} memory models were defined in the specification. Moreover, most modern high-level programming languages rely on relaxed memory model. Thus, the memory model for Java that is based on the \textit{happens-before} principle~\cite{lamport1978time} was introduced in J2SE 5.0 in 2004~\cite{manson2005java}; the C++11 standard~\cite{?} has introduced the set of hardware-independent synchronisation fences and atomic operations, whenever the C++17 memory model is based on the relation \textit{strongly happens-before}.
%batty2011mathematizing
Thirdly, important research direction targets the problem of verifying (or at least finding bugs in) existing software systems with respect to WMMs. Perhaps, the most notable work in this field is the project for defining the Linux kernel memory model, which is being actively developing these days~\cite{kernel1}.
%Distributed databases <also need the wmm, see transaction consistency~\cite{bailis2013highly}>

%experiments with kernel?

@@

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{Project web page: \url{diy.inria.fr}} is a sofware suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy}, \tool{diycross} and \tool{diyone}, the litmus tests concrete executor \tool{litmus} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd} that implements reachability analysis for exploring states reachable under specified~WMM.

All the \tool{diy} tools work only with single memory model, however, in real life we face serious engineering problems involving necessity to model more than one execution environment. One of these problems is the \textit{portability} of the program from one hardware architecture to another. A program written in a high-level language is then compiled for different hardware. Even if all the compiler optimisations were disabled (which is rare case nowadays), the behaviour of two compiled versions of the same program may differ due to differences between hardware memory models.
As the result, a program compiled under the platforms $T$ can reach states that are unreachable on the platform $S$, which is a \textit{portability bug} from the source platform $S$ to the target platform $T$~\cite{Porthos17}.

### 2018-04-18 2_wmm.tex
Cleanup intro and wmm chapters.

\chapter{Weak Memory Models}
\label{ch:wmm}

%"Semantics and verification under weak memory models have been the subject of study at least since 2007." [PORTHOS]
A \textit{weak memory model} is a set of predicative constraints over so-called \textit{derived relations}, which defines the set of possible executions of a concurrent program.
Usually, these constraints include requirements of the reflexivity or acyclicity of derived relations.
The study of formalisation weak memory models for different architectures has been rapidly developed over last decade.

Firstly, the research of WMM aimes to formalise the weak memory models and provide systematic, sound and complete formal approach of defining WMMs in order to be able to verify systems with respect to them.
A generic framework of program analysis with respect to weak memory models was presented in 2010~\cite{alglave2010shared}. Similar approach is used \Porthos as it is described
% in Section~\cite{ch:wmm:event}. 
further in current Chapter.

Secondly, researchers work on extracting the WMMs of hardware architectures from existing implementations of from their specifications, that are written in natural language and thus suffer from ambiguities and incompletenesses. Over last decade the memory models have been extracted for most mainstream multiprocessor architectures, such as x86-TSO (\textit{Total Store Order}) model for x86 architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding}, for Alpha~\cite{?} and Sparc~\cite{?} memory models were defined in the specification. Moreover, most modern high-level programming languages rely on relaxed memory model. Thus, the memory model for Java that is based on the \textit{happens-before} principle~\cite{lamport1978time} was introduced in J2SE 5.0 in 2004~\cite{manson2005java}; the C++11 standard~\cite{?} has introduced the set of hardware-independent synchronisation fences and atomic operations, whenever the C++17 memory model is based on the relation \textit{strongly happens-before}.
%batty2011mathematizing

Thirdly, important research direction targets the problem of verifying (or at least finding bugs in) existing software systems with respect to weak memory models. Perhaps, the most notable work in this field is the project for defining the Linux kernel memory model, which is being actively developing these days~\cite{kernel1}.
%Distributed databases <also need the wmm, see transaction consistency~\cite{bailis2013highly}>

%experiments with kernel?

@@

The classical approach to model the concurrent programs is to use the \textit{global time}, a single order of interleavings of all actions happened in different threads. However these models are easy to understand, it may be hard to consider \textit{all} possible states, number of which is exponentially large. Another way to do this is to to use non-deterministic computation-centric models defined in~\cite{fri97}, one of which represents the program as the graph of \textit{memory events}. The idea in this class of models is based on the fact that the behaviour of a concurrent system is defined only by the interleavings of shared-memory operations, while being independant from the order of local computation events. These models may be further restricted by constraints of a weak memory model, adding \textit{relations} to the memory events.
%Perhaps, the most convenient way to model the non-deterministic properties of concurrent programs is to use the program representation based on \textit{memory events}.
%In order to model the non-deterministic properties of concurrent programs

The event-based program model represents the directed graph (the \textit{event-flow graph}), where vertices represent \textit{events}, and edges represent \textit{relations} over the events. An event is something, that, after being executed, changes the state of an abstract machine executing the concurrent program. 
An \textit{execution} (trace, run) of a given program is an ordered set of events. The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph. 
An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}. An execution to be checked on validity is called the \textit{candidate execution}.

Below we describe some basic types of events and relations.

@@

The \textit{data-flow} of a program is defined by \textit{communication relations}: the \rf-relation $\subset~\mathbb{W}~\times~\mathbb{R}$ (\textit{read-from} relation) that maps a write to a read reading its value, the \co-relation $\subset~\mathbb{W}~\times~\mathbb{W}$ (\textit{coherence order}, sometimes called \texttt{ws}-relation for \textit{write serialisation}) defines the total order on writes to the same location across all processes, and the \fr-relation $\subset~\mathbb{R}~\times~\mathbb{W}$ (\textit{from-read order}) that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

Figure~\ref{simple_wmm_x86_pic} illustrates the candidate execution for the Example~\ref{simple_wmm_x86}, that reaches the state \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)} within x86-TSO memory model as it forbids cycles over \po$\cup$\rf (union) relation (the picture is generated by the \texttt{herd7} tool, version 7.47). 
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

The semantics of a concurrent program is represented by the set of allowed executions.
An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$ (\textit{data-flow} of a program)~\cite{alglave2010shared}. As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program as a whole.

>>>>

\chapter{Memory Model-Awared Program Analysis}
\label{ch:wmm}

In general, analysis of concurrent programs with respect to axiomatic memory models require firstly to capture the control-flow of a concurrent program, after which the set of all possible data-flows (called \textit{candidate executions}) is computed as the set of possible communications between threads of the program. After that, the \textit{weak memory model} specification constraints this set of candidate executions and decides which execution is valid~\cite{alglave2014herding}.
More specifically, we use the event-based program representation (see Section~\ref{ch:wmm:model:events}) as a natural way to capture the semantics of a concurrent program. Therefore, a candidate execution is defined as a choice of events to be executed. The constraints over events imposed by both program and memory model are called \textit{relations} (see Section~\ref{ch:wmm:model:relations})

%More presice description of x86-TSO memory model is given in Chapter~\ref{ch:wmm:x86}.

@@

The classical approach to model concurrent programs is to use the \textit{global time}, a single order of interleavings of all actions happened in different threads. However these models are easy to understand, it may be hard to consider \textit{all} possible states, number of which is exponentially large. Another way to do this is to to use non-deterministic computation-centric models defined in~\cite{fri97}, one of which represents the program as the graph of \textit{memory events}. The idea in this class of models is based on the fact that the behaviour of a concurrent system is defined only by the interleavings of shared-memory operations, while being independent from the order of local computation events.
%These models may be further restricted by constraints of a weak memory model, adding \textit{relations} to the memory events.
%Perhaps, the most convenient way to model the non-deterministic properties of concurrent programs is to use the program representation based on \textit{memory events}.
%In order to model the non-deterministic properties of concurrent programs

The event-based program model represents the directed graph (the \textit{event-flow graph}), where vertices represent \textit{events} (simple low-level instructions), and edges represent \textit{relations} over the events.
Below we describe some basic types of events and relations.

@@

The \textit{data-flow} of a program is defined by \textit{communication relations}: the \rf-relation $\subset~\mathbb{W}~\times~\mathbb{R}$ (\textit{read-from} relation) that maps a write event to the read event that reads its value, the \co-relation $\subset~\mathbb{W}~\times~\mathbb{W}$ (\textit{coherence order}, sometimes called \texttt{ws}-relation for \textit{write serialisation}) defines the total order on writes to the same location across all processes, and the \fr-relation $\subset~\mathbb{R}~\times~\mathbb{W}$ (\textit{from-read order}) that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

The semantics of a concurrent program is represented by the set of allowed executions.
A \textit{candidate execution} (trace, run) of a program is an ordered set of events defined by \po and \rf relations and set of final writes to a given memory location~\cite{alglave2014herding}.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.

%associated with the instructions of the program
%the path in the event-flow graph. 
An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}. 
An execution to be checked on validity is called the \textit{candidate execution}.
An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$~\cite{alglave2010shared}.
As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program as a whole.

The Figure~\ref{simple_wmm_x86_pic} illustrates the candidate execution for the Example~\ref{simple_wmm_x86}, that reaches the state \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)} within x86-TSO memory model as it forbids cycles over \po$\cup$\rf (union) relation (the picture is generated by the \texttt{herd7} tool, version 7.47). 
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@
